---
title: "Bed Occupancy Forecasting Workflow"
author: "Bedcast Package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bed Occupancy Forecasting Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6
)
```

# Introduction

The `bedcast` package provides tools for forecasting bed occupancy during epidemics using Bayesian nowcasting methods. This vignette demonstrates the complete workflow for analyzing epidemiological data and generating forecasts of ETU (Ebola Treatment Unit) and isolation bed occupancy.

# Workflow Overview

The bedcast workflow consists of several key steps:

1. **Data Import**: Import and process linelist, alerts, and occupancy data
2. **Delay Estimation**: Calculate delay distributions between key events
3. **Data Merging**: Combine all data sources into a time series format
4. **Model Fitting**: Fit a Bayesian Stan model for nowcasting and forecasting
5. **Visualization**: Create comprehensive plots of results

# Step-by-Step Workflow

## 1. Load Required Packages

```{r load-packages, eval = FALSE}
# Load the bedcast package and dependencies
library(bedcast)
library(tidyverse)
library(rstan)
library(here)
```

## 2. Import Data

The workflow begins by importing data from various sources:

### Import Linelist Data
```{r import-linelist, eval = FALSE}
# Import real linelist data
ll <- get_real_linelist()
```

The `get_real_linelist()` function imports and processes epidemiological linelist data, handling various date formats and cleaning column names.

### Import Alerts Data
```{r import-alerts, eval = FALSE}
# Import daily alerts data
alerts <- get_alerts()
```

The `get_alerts()` function processes daily alerts data, converting from wide format (districts as columns) to long format suitable for analysis.

### Import Occupancy Data
```{r import-occupancy, eval = FALSE}
# Import ETU and isolation occupancy data
etu_iso <- get_etu_iso()
```

The `get_etu_iso()` function imports daily bed occupancy data for treatment and isolation facilities.

### Import Incidence Data
```{r import-incidence, eval = FALSE}
# Import incidence data (optional - can be calculated from linelist)
incidence <- get_incidence()
```

## 3. Calculate Delay Distributions

Delay distributions are crucial for understanding the time between different epidemiological events:

```{r calculate-delays, eval = FALSE}
# Calculate delay distributions from linelist data
delays <- get_delays(ll)
```

The `get_delays()` function estimates delay distributions for:
- Onset to ETU admission
- ETU admission to survival/death
- Onset to contact isolation
- Lab turnaround time
- Onset to case confirmation

## 4. Merge Data Sources

Combine all data sources into a single time series:

```{r merge-data, eval = FALSE}
# Merge all data sources
data <- merge_data(ll, alerts, etu_iso, delays, incidence)
```

The `merge_data()` function:
- Calculates daily incidence from linelist data
- Aggregates alerts by date
- Fills gaps in the time series
- Creates case inflation factors for reporting delays
- Converts to tsibble format for time series analysis

## 5. Compile and Fit Stan Model

### Compile the Stan Model
```{r compile-model, eval = FALSE}
# Compile the Stan model
model <- stan_model(here("inst/stan/occupancy_model_nowcast.stan"))
```

### Fit the Model
```{r fit-model, eval = FALSE}
# Fit the model with custom priors
results <- fit_stan(
  data, model,
  # Delay priors (log-mean, log-sd)
  prior_onset_to_etu = delays$onset_to_etu,
  prior_etu_to_survival = delays$etu_to_survival,
  prior_etu_to_death = delays$etu_to_death,
  prior_onset_to_iso = delays$onset_to_iso,
  prior_iso_to_release = c(log(3), 0.5, 1.5, 0.5),
  
  # Case fatality rate prior (logit scale)
  prior_cfr = c(qlogis(0.40), 0.3),
  
  # Isolation parameters
  prior_prop_iso = c(qlogis(0.8), 0.8),
  
  # Alert parameters
  prior_alerts_background = c(log(50), 0.4),
  prior_alerts_per_case = c(log(10), 0.4),
  
  # Model settings
  days_ahead = 7*6,  # 6 weeks ahead
  n_iter = 1e3,      # 1000 iterations
  n_chains = 3,      # 3 chains
  n_knots = 20       # 20 knots for growth rate spline
)
```

The `fit_stan()` function fits a hierarchical Bayesian model that:
- Models case incidence using spline-based growth rates
- Accounts for reporting delays using delay distributions
- Links cases to bed occupancy through admission/discharge processes
- Models alerts as a function of cases and background noise
- Provides nowcasts and forecasts with uncertainty quantification

## 6. Visualize Results

### Model Fit Visualization
```{r visualize-fit, eval = FALSE}
# Create comprehensive model fit visualization
p_fit <- vis_stan_fit(results, ylim_fact = 1.5)
p_fit %>% save_plot("model_fit.png", height = 20)
```

The `vis_stan_fit()` function creates a comprehensive visualization showing:
- Observed data as bar charts
- Nowcasts and forecasts with credible intervals
- Separate panels for cases, ETU occupancy, alerts, and isolation occupancy
- Clear distinction between observed and projected data

### Parameter Distributions
```{r visualize-parameters, eval = FALSE}
# Visualize parameter distributions and priors
p_params <- vis_parameters(results, base_size = 14)
p_params %>% save_plot("parameters.png", width = 21, height = 15.3)
```

The `vis_parameters()` function shows:
- Posterior parameter distributions (informed by data)
- Prior distributions (initial assumptions)
- Comparison of how much the data informed each parameter

### Growth Rate Visualization
```{r visualize-growth, eval = FALSE}
# Visualize growth rate over time
p_growth <- vis_growth_rate(results, base_size = 14)
p_growth %>% save_plot("growth_rate.png")
```

The `vis_growth_rate()` function shows:
- Daily growth rate estimates over time
- Uncertainty bounds from MCMC samples
- Reference line at zero growth

## 7. Extract and Analyze Results

### Summarize Parameter Estimates
```{r summarize-results, eval = FALSE}
# Get case estimates with credible intervals
cases <- summarise_stan("log_cases_inflated", results, 
                       probs = c(0.025, 0.5, 0.975))

# Get ETU occupancy estimates
etu_occ <- summarise_stan("etu_modelled", results)

# Get case fatality rate
cfr <- summarise_stan("cfr", results)
```

### Extract Raw Samples
```{r extract-samples, eval = FALSE}
# Extract raw MCMC samples for custom analysis
cfr_samples <- extract_stan("cfr", results)
growth_samples <- extract_stan("growth_rate_vec", results)
```

# Key Features

## Bayesian Nowcasting
The package implements Bayesian nowcasting methods that:
- Account for reporting delays in case data
- Provide uncertainty quantification
- Generate both nowcasts and forecasts

## Flexible Delay Modeling
Delay distributions are estimated from data and can be:
- Customized with different prior assumptions
- Updated as new data becomes available
- Used to inform case inflation factors

## Comprehensive Visualization
The package provides multiple visualization functions:
- Model fit plots with credible intervals
- Parameter distribution comparisons
- Growth rate time series
- Customizable plot saving

## Stan Integration
The package leverages Stan for:
- Efficient MCMC sampling
- Flexible model specification
- High-quality posterior inference

# Best Practices

## Data Quality
- Ensure consistent date formats across data sources
- Check for missing values and outliers
- Validate delay distributions make epidemiological sense

## Model Tuning
- Adjust priors based on domain knowledge
- Increase MCMC iterations for better convergence
- Use multiple chains to assess convergence

## Visualization
- Save plots in high resolution for publications
- Customize plot aesthetics for your needs
- Use appropriate credible interval levels

# Conclusion

The `bedcast` package provides a comprehensive toolkit for bed occupancy forecasting during epidemics. The workflow is designed to be flexible and adaptable to different disease contexts while maintaining rigorous statistical methods.

For more information on specific functions, see the package documentation and help files.
