---
title: "Bed Occupancy Forecasting Workflow"
author: "Finlay Campbell"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bed Occupancy Forecasting Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5
)
```

## Introduction

The **bedcaster** package forecasts healthcare demand during epidemics by fitting a single Bayesian model to daily surveillance data. The model nowcasts and projects:

- reported cases and deaths (with reporting-delay adjustment),
- ETU occupancy,
- community alerts, and
- isolation bed occupancy.

This vignette outlines the workflow from a prepared data frame through fitting, summarisation, and visualisation.

## Data preparation

`fit_bedcaster()` expects a data frame with columns `date`, `cases`, `deaths`, `etu`, `alerts`, and `iso`. Build this from your linelist, bed-management system, and alert database. Counts can be `NA` on days when a series was not recorded.

```{r example-data, eval = FALSE}
library(bedcaster)

df <- data.frame(
  date = seq(as.Date("2024-01-01"), as.Date("2024-03-31"), by = "day"),
  cases = rpois(91, 8),
  deaths = rpois(91, 1),
  etu = rpois(91, 12),
  alerts = rpois(91, 20),
  iso = rpois(91, 15)
)

as_of <- max(df$date)
```

In practice you will merge linelist-derived incidence, ETU/isolation occupancy, and alert time series before calling the model. The package does not ship outbreak-specific import helpers; data preparation stays in your analysis pipeline.

## Fitting the model

```{r fit, eval = FALSE}
bedcast <- fit_bedcaster(
  df,
  as_of = as_of,
  n_proj = 28,
  n_iter = 500,
  n_chains = 2,
  n_cores = 2
)

print(bedcast)
```

Key arguments:

- **`as_of`** — snapshot date used to compute reporting-delay weights.
- **`n_proj`** — projection length in days beyond the fit period.
- **`extrapolate_growthrate`** — if `FALSE` (default), projected growth rates stay at the last estimated value; if `TRUE`, growth is extrapolated with a logistic asymptote.
- **Delay priors** — length-4 vectors for each delay (`meanlog.mean`, `meanlog.sd`, `sdlog.mean`, `sdlog.sd`); see `?fit_bedcaster`.

```{r priors, eval = FALSE}
delay_prior <- c(
  meanlog.mean = log(5), meanlog.sd = 0.5,
  sdlog.mean = log(2), sdlog.sd = 0.5
)

bedcast <- fit_bedcaster(
  df,
  as_of = as_of,
  prior_onset_to_etu = delay_prior,
  prior_cfr = c(qlogis(0.3), 0.25)
)
```

Increase `n_iter` and `n_chains` for production runs and check convergence with `rstan::check_hmc_diagnostics(bedcast$fit)`.

## Summaries and extraction

```{r summary, eval = FALSE}
# Quantiles by date for a vector parameter
summary(bedcast, "cases_nowcast_sim", alpha = 0.95)

# Scalar parameter
summary(bedcast, "cfr")

# Long-format MCMC draws
extract(bedcast, "cfr")
```

## Visualisation

```{r plots, eval = FALSE}
plot(bedcast, type = "fit")
plot(bedcast, type = "parameters")
plot(bedcast, type = "growthrate")
```

`vis_bedcast_fit()` shows reported bars, nowcast/forecast ribbons, and median lines for each outcome. `vis_bedcast_parameters()` overlays prior (grey) and posterior (green) densities. `vis_bedcast_growthrate()` plots the spline-based daily growth rate with credible intervals.

## Best practices

1. **Align `as_of` with your reporting snapshot** so delay weights match the data you present to decision-makers.
2. **Set priors from epidemiological knowledge** — delays and CFR should reflect the outbreak context.
3. **Use enough MCMC iterations** for stable intervals before operational use.
4. **Review all outcomes** — poor fit on ETU or isolation may indicate data issues even when cases look reasonable.

## Further reading

- Package README: [github.com/finlaycampbell/bedcaster](https://github.com/finlaycampbell/bedcaster)
- Function reference: `?fit_bedcaster`, `?vis_bedcast_fit`
